---
title: "Chemogenomic profiling of diverse
Saccharomyces cerevisiae strains using the novel CRISPR-Cas9 markerless
barcoded library BarMix"
author: "Jackson Moore, Marjan Barazandeh, Corey Nislow, Vivien Measday"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
params:
  run_analysis: false
---

This code accompanies the manuscript *Chemogenomic profiling of diverse
Saccharomyces cerevisiae strains using the novel CRISPR-Cas9 markerless
barcoded library BarMix*.

## Overview and requirements

The workflow creates a barcode reference, maps paired-end sequencing reads,
counts barcode overlaps, estimates strain fitness, and generates exploratory
plots.

Place this document in the project directory alongside:

- `Additional File 1 - Supplemental Tables.xlsx`: strain/barcode/clade information
  in sheet 1 and compound information in sheet 2
- `BarSeq_Directory/`: paired FASTQ files

Required command-line tools: BWA, SAMtools, and BEDTools. Required R packages:
`readxl`, `tidyverse`, `phylotools`, `ggsignif`, `pheatmap`, `Hmisc`,
`tidygraph`, `ggraph`, `factoextra`, `knitr`, and `rmarkdown`.

## Load R libraries
```{r}
library(readxl)
library(tidyverse)
library(phylotools)
library(ggpubr)
library(pheatmap)
library(viridis)
library(RColorBrewer)
```

## R: Create the barcode FASTA reference

The following code generates an indexed bed file to be used for mapping barcode sequences to their corresponding strain identity

#R: Create a FASTA file of barcodes
```{r create-fasta}
#Read in Table S1 containing barcode amplicon sequences and corresponding strains
raw_dat <- read_xlsx('Additional File 1 - Supplemental Tables.xlsx', sheet = 1, skip = 2)
dat <- raw_dat %>% 
  select(Strain, `Barcode`) 
dat <- dat %>% 
  rename("seq.name" = Strain) %>% 
  rename("seq.text" = Barcode)
#Convert to fasta file
dat2fasta(dat, outfile = "Amplicon_Database.fa")
```

## Terminal (Bash): Index the reference and generate bed file

```{bash index-fasta, eval = FALSE}
#Index fasta file
bwa index Amplicon_Database.fa
#Convert fasta to bed format:
# Use awk to process the FASTA file and generate a BED file
awk -v OFS='\t' '/^>/ {if(seqname){print seqname, 1, seqlen, ".", ".", "+"}; seqname=substr($1,2); seqlen=0; next} {seqlen+=length($0)} END{print seqname, 1, seqlen, ".", ".", "+"}' Amplicon_Database.fa > Amplicon_Database.bed
```

## Terminal (Bash): Map barcode reads to corresponding strains

```{bash bwa-mapping, eval = FALSE}
#The BarSeq directory should contain all raw fastq files obtained from Bar-seq experiments
for file in BarSeq_Directory/*_1.fastq
do
f=$(basename "$file")
prefix="${f%_1.fastq}"  # Remove _1.fastq to get the sample name prefix
paired_end_1="BarSeq_Directory/${prefix}_1.fastq"
paired_end_2="BarSeq_Directory/${prefix}_2.fastq"
barcode_database=Amplicon_Database.fa

echo "Running bwa mem on $paired_end_1 and $paired_end_2"

#Perform bwa mapping to generate bam files as output
bwa mem -t 4 $barcode_database ${paired_end_1} ${paired_end_2} >  BarSeq_Directory/${prefix}.sam
samtools view -b -o BarSeq_Directory/${prefix}.bam  BarSeq_Directory/${prefix}.sam
done
```

## Terminal (Bash): Sort, filter, and count barcode depth per strain

```{bash barcoding-counting, eval = FALSE}
for file in BarSeq_Directory/*.bam
do
f=$(basename "$file")
prefix="${f%.bam}"  # Remove _1.fastq to get the sample name prefix
bed_file=Amplicon_Database.bed

echo "Running bwa mem on $prefix"

#Sort bam files
samtools sort -O bam -o BarSeq_Directory/${prefix}.sorted.bam -T BarSeq_Directory${prefix}_tmp -@ 4 BarSeq_Directory/${prefix}.bam
#Filter bam files with quality mapping score 30
samtools view -q 30 -h -b -o BarSeq_Directory/${prefix}.q.bam BarSeq_Directory/${prefix}.sorted.bam
#index bam files
samtools index BarSeq_Directory/${prefix}.q.bam
#Count number of reads per barcode
coverageBed -a $bed_file -b BarSeq_Directory/${prefix}.q.bam -s > BarSeq_Directory/${prefix}.q_counts.txt
done
```

## R: Assemble the raw-count matrix

```{r}
#Function for reading bwa counts
read_bwa_counts = function(cf, bwa_out_dir = NULL){
  # Extract the sample name  of the filename
  s = paste(unlist(strsplit(x = cf, split = "_"))[3:4], collapse = "_")
  f = file.path(bwa_out_dir, cf)
  
  counts_f = read.table(file = f, header = F, sep = "\t", quote = "", comment.char = "", stringsAsFactors = F)
  #Extract sample name and count information
  counts_f1 = counts_f %>% select(V1, V7)
  
  counts2 = counts_f1 %>% rename(Strain = V1, counts = V7) 
  
  counts4 = counts2 %>% as.data.frame()
  #Name column with sample name
  colnames(counts4)[2] = s
  
  return(counts4)
}

#Run function across all Bar-seq data
counts_dir = file.path("BarSeq_Directory/")
count_files = dir(path = "BarSeq_Directory/", pattern = "*.counts.txt")
bwa_counts = NULL
for(f in count_files){
  c = read_bwa_counts(cf = f , bwa_out_dir = counts_dir)
  if(is.null(bwa_counts)){
    bwa_counts = c
  } else {
    bwa_counts = bwa_counts %>% left_join(c)
  }
}

#Save csv file of raw counts
write_csv(bwa_counts, 'Raw_Counts.csv')
```

## R: Assess correlations between samples

```{r}
#Read in raw counts data
bwa_counts <- read_csv('Raw_Counts.csv')

bwa_counts_replicates <- bwa_counts

# Exclude the first column which contains Strain names
bwa_counts_replicates2 <- bwa_counts_replicates[, -1]

# Calculate pairwise correlation
cor_matrix <- cor(bwa_counts_replicates2)

# Convert correlation matrix to a dataframe
cor_df <- as.data.frame(as.table(cor_matrix))

# Rename columns for clarity
names(cor_df) <- c("Replicate_Condition_1", "Replicate_Condition_2", "Correlation")

#Visualize data via ggplot
ggplot(cor_df, aes(Replicate_Condition_1, Replicate_Condition_2)) +
  geom_tile(aes(fill = Correlation)) +
  geom_text(aes(label = round(Correlation, 2)), size = 1.5) +
  scale_fill_gradient(low = "red", high = "green", limits=c(0,1)) +
  xlab("Sample") + 
  ylab("Sample") +
    theme(axis.text.x = element_text(angle = 90, size = 4), axis.text.y = element_text(size = 4, face = "bold"), axis.title.x = element_text(size = 20, face = "bold"), axis.title.y = element_text(size = 20, face = "bold"), legend.text = element_text(size = 14), legend.title = element_text(size = 14, face = "bold"))
```

## R: Normalize within a batch
Here, mean read depths are normalized across samples within a single batch. Normalized read depths of each Strain in drug-treated conditions are further normalized to their corresponding read depth in control conditions (i.e., YPD, DMSO, YPEG) to determine Strain fitness values. In this case, read depths are normalized within batch, and not across batches 

```{r}
#Read in raw counts data
bwa_counts <- read_csv('Raw_Counts.csv')

#Pivot to long format. Note, change the cols option to reflect the number of columns in the batch dataset
bwa_counts_normalized <- bwa_counts %>% 
    pivot_longer(cols = 2:137, names_to = "Pool", values_to = "Read_Depth") 

#Create column indicating which drug treatment each replicate belongs to
bwa_counts_normalized <- bwa_counts_normalized %>%
  mutate(Condition = str_extract(Pool, "^[^_]+"))

#Calculate mean of each strain/condition for normalized read depth and remove duplicates
bwa_counts_normalized <- bwa_counts_normalized %>% 
  group_by(Condition, Strain) %>% 
  mutate(Mean_Depth_Strain_Condition = mean(Read_Depth))

bwa_counts_normalized <- bwa_counts_normalized %>% 
  select(Strain, Condition, Mean_Depth_Strain_Condition) %>% 
  distinct()

#Filter out strains with low read depth in control condition (e.g., YPD). Note: Change control condition below to reflect the specific sample control within batch (i.e., YPD/DMSO/YPEG)
bwa_counts_normalized <- bwa_counts_normalized %>% 
  mutate(Mean_Depth_Strain_Condition = ifelse(Condition == "YPD" & Mean_Depth_Strain_Condition < 20, NA, Mean_Depth_Strain_Condition))
bwa_counts_normalized <- bwa_counts_normalized %>% 
  drop_na()

#Normalize read depth of each drug treatment to mean read depth of control across all strains (e.g., YPD). Note: Change control condition below to reflect the specific sample control within batch (i.e., YPD/DMSO/YPEG)
bwa_counts_normalized <- bwa_counts_normalized %>% 
  ungroup() %>% 
  group_by(Condition) %>% 
  mutate(Mean_Depth_Condition = mean(Mean_Depth_Strain_Condition))

bwa_counts_normalized <- bwa_counts_normalized %>% 
 ungroup() %>% 
 mutate(Normalization_Factor_YPD_overall = Mean_Depth_Condition[Condition == "YPD"][1]/Mean_Depth_Condition) %>% 
 mutate(Normalized_Depth_YPD_Overall = Normalization_Factor_YPD_overall * Mean_Depth_Strain_Condition)

#Divide read depth of each strain by its read depth in control condition to determine strain fitness factor. Note: Change control condition below to reflect the specific sample control within batch (i.e., YPD/DMSO/YPEG)
bwa_counts_normalized <- bwa_counts_normalized %>% 
  ungroup() %>% 
  group_by(Strain) %>% 
  #When normalizing to whole YPD counts
  mutate(Strain_Fitness = Normalized_Depth_YPD_Overall/Normalized_Depth_YPD_Overall[Condition == "YPD"][1]) 

#Select and remove duplicates
bwa_counts_normalized <- bwa_counts_normalized %>% 
  select(Strain, Condition, Strain_Fitness)
bwa_counts_normalized <- bwa_counts_normalized %>% 
  distinct()
bwa_counts_normalized <- bwa_counts_normalized %>% 
  drop_na()

#Save file
write_csv(bwa_counts_normalized, 'Normalized_counts.csv')
```

## R: Scale samples across batches and estimate fitness
Here, mean read depths are normalized across samples and across batches. Each batch is normalized by determining the median read depth of each control condition across batches. Then, normalized read depths of each strain in drug-treated conditions are further normalized to their corresponding read depth in control conditions (i.e., YPD, DMSO, YPEG) to determine strain fitness values. In this case, read depths are normalized across batches to reduce batch-effect bias. 

```{r}
#Read in raw counts information. This file should contain all samples across all batches
bwa_counts <- read_csv('raw_counts.csv')

# Calculate the median for each sample
medians <- apply(bwa_counts[, -1], 2, median)

# Normalize each sample to the same median (e.g., median of medians)
target_median <- median(medians)

df_normalized_median <- bwa_counts %>%
  mutate(across(-Strain, ~ . * (target_median / median(.))))

#Sanity check for medians
medians <- apply(df_normalized_median[, -1], 2, median)

#Pivot to long format. Note, change cols option to reflect the number of columns in dataset
bwa_counts_normalized <- df_normalized_total %>% 
    pivot_longer(cols = 2:158, names_to = "Pool", values_to = "Read_Depth")

#Create column indicating which treatment each sample belongs to
bwa_counts_normalized <- bwa_counts_normalized %>%
  mutate(Condition = str_extract(Pool, "^[^_]+")) 

#Calculate mean of each strain/condition for normalized read depth and remove duplicates
bwa_counts_normalized <- bwa_counts_normalized %>% 
  group_by(Condition, Strain) %>% 
  mutate(Mean_Depth_Strain_Condition = mean(Read_Depth))

bwa_counts_normalized <- bwa_counts_normalized %>% 
  select(Strain, Condition, Mean_Depth_Strain_Condition) %>% 
  distinct()

#Normalize read depth to YPD, DMSO and T = 0
bwa_counts_normalized <- bwa_counts_normalized %>% 
  ungroup() %>% 
  group_by(Strain) %>% 
  #When normalizing to whole YPD counts
  mutate(Normalized_depth_YPD = Mean_Depth_Strain_Condition/Mean_Depth_Strain_Condition[Condition == "YPD"][1]) %>% 
  mutate(Normalized_depth_SCD = Mean_Depth_Strain_Condition/Mean_Depth_Strain_Condition[Condition == "SCD"][1]) %>% 
  mutate(Normalized_depth_YPD35 = Mean_Depth_Strain_Condition/Mean_Depth_Strain_Condition[Condition == "YPD35"][1]) %>% 
  mutate(Normalized_depth_YPEGD = Mean_Depth_Strain_Condition/Mean_Depth_Strain_Condition[Condition == "YPEGD"][1]) %>% 
  mutate(Normalized_depth_T = Mean_Depth_Strain_Condition/Mean_Depth_Strain_Condition[Condition == "T"][1]) %>% 
  mutate(Normalized_depth_YPDNoDMSO = Mean_Depth_Strain_Condition/Mean_Depth_Strain_Condition[Condition == "YPDNoDMSO"][1]) 

#Select and remove duplicates
bwa_counts_normalized <- bwa_counts_normalized %>% 
  select(Strain, Condition, Normalized_depth_YPD, Normalized_depth_YPEGD, Normalized_depth_SCD, Normalized_depth_YPD35, Normalized_depth_T, Normalized_depth_YPDNoDMSO)
bwa_counts_normalized <- bwa_counts_normalized %>% 
  distinct()
bwa_counts_normalized <- bwa_counts_normalized %>% 
  drop_na()

#Normalize each treatment to their corresponding control condition and determine strain fitness values

bwa_counts_normalized_condense <- bwa_counts_normalized %>%
  mutate(Normalized_Fitness = ifelse(Condition == "ANT" | Condition == "OLI", Normalized_depth_YPEGD,
                                     ifelse(Condition == "YPDNoDMSO" | Condition == "YPD", Normalized_depth_T, 
                                            ifelse(Condition == "MMS0.0125" | Condition == "MMS0.025" | Condition == "YPEG", Normalized_depth_YPDNoDMSO,
                                                   ifelse(Condition == "ETH" | Condition == "HYD" | Condition == "CAF", Normalized_depth_YPD35, Normalized_depth_YPD)))))

bwa_counts_normalized_condense <- bwa_counts_normalized %>%
  mutate(Normalized_Fitness = ifelse(Condition == "ANT" | Condition == "OLI", Normalized_depth_YPEGD,
                                     ifelse(Condition == "YPDNoDMSO", Normalized_depth_T, 
                                            ifelse(Condition == "YPEG", Normalized_depth_YPDNoDMSO,
                                                   ifelse(Condition == "ETH" | Condition == "HYD" | Condition == "CAF", Normalized_depth_YPD35, Normalized_depth_YPD)))))

bwa_counts_normalized_condense <- bwa_counts_normalized %>%
  mutate(Normalized_Fitness = ifelse(Condition == "ANT" | Condition == "OLI", Normalized_depth_YPEGD,
                                     ifelse(Condition == "YPDNoDMSO" | Condition == "YPD", Normalized_depth_T, 
                                            ifelse(Condition == "YPEG", Normalized_depth_YPDNoDMSO,Normalized_depth_YPD))))
                                                   
bwa_counts_normalized_condense <- bwa_counts_normalized_condense %>% 
  select(Strain, Condition, Normalized_Fitness)

#Save file
write_csv(bwa_counts_normalized_condense, 'Batch_Normalized_ReadDepth_counts.csv')
```

## R: Add phylogenetic annotations and plot one condition

```{r}
#Read in normalized counts data
bwa_counts_normalized <- read_csv('Normalized_counts.csv')

#Read in phylogenetic information
phylo <- read_xlsx('Additional File 1 - Supplemental Tables.xlsx', sheet = 1, skip 2)

phylo <- phylo %>% 
  select(Strain, `Phylogenetic Clade`)

#Add phylogenetic information to counts
bwa_counts_normalized_phylo <- bwa_counts_normalized %>% 
  left_join(phylo, by = "Strain")
```

## R: Prepare Strain-by-condition matrix

```{r}
#Identify outliers
#Find outliers 
is_outlier <- function(x) {
  return(x < quantile(x, 0.25) - 1.5 * IQR(x) | x > quantile(x, 0.75) + 1.5 * IQR(x))
}

#Filter dataset for specific drug (replace X with compound name)
bwa_counts_normalized_drug <- bwa_counts_normalized_phylo %>%
  filter(Condition == "X") %>% 
  drop_na()

bwa_counts_normalized_drug <- bwa_counts_normalized_drug %>%
  ungroup() %>% 
  group_by(`Phylogenetic Clade`) %>% 
  mutate('Outlier' = ifelse(is_outlier(Strain_Fitness), Strain, (NA)))

# Convert 'group' column to a factor with specified levels
bwa_counts_normalized_drug$`Phylogenetic Clade` <- factor(bwa_counts_normalized_drug$`Phylogenetic Clade`, levels = c(
   "CHN", 
  "Far East Asia", 
  "Ecuadorian", 
  "Mediterranean Oak", 
  "Transpacific Oak", 
  "Pacific West Coast Wine", 
  "Wine/European",
  "Prise de Mousse/Vin7", 
  "Laboratory", 
  "Beer 1/Mixed Origin",
  "Mosaic"
))

#GGplot to visualize
ggplot(bwa_counts_normalized_drug) +
  aes(x = `Phylogenetic Clade`, y = Normalized_depth_YPD, fill = `Phylogenetic Clade`) +
  geom_boxplot() +
  theme_bw() +
  scale_fill_manual(
    values = c(
     "Wine/European" = "#E07082",
      "Pacific West Coast Wine" = "#61d04f",
      "Transpacific Oak" = "#8F6249",
      "Mosaic" = "#c7c7c7",
      "CHN" = "#8c32bb",
      "Beer 1/Mixed Origin" = "#f47c11",
      "Ecuadorian" = "#85C9F1",
      "Far East Asia" = "#6666FF",
      "Mediterranean Oak" = "#b28a6e",
      "Prise de Mousse/Vin7" = "#B24A5A",
      "Laboratory" = "#FFB3FF")) +
  ylab("Normalized Read Depth (Drug - DMSO)") + 
  theme_classic() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1, size = 19, face = "bold"), axis.text.y = element_text(size = 19, face = "bold"), axis.title.x = element_blank(), axis.title.y = element_text(size = 19, face = "bold"), panel.grid.major = element_line(size = 0.5, linetype = 'dashed', color = 'grey'), panel.grid.minor = element_blank()) +
    geom_signif(comparisons = list(c("Transpacific Oak", "Pacific West Coast Wine"),c("Transpacific Oak", "Wine/European"), c("Pacific West Coast Wine", "Wine/European")), test = "t.test", map_signif_level = T)
```

## R: Plot Strain-fitness heatmaps

```{r}
#Read in normalized counts data
bwa_counts_normalized_condense <- read_csv('Batch_Normalized_ReadDepth_counts.csv')

phenotype <- bwa_counts_normalized_condense %>% 
    rename("Compound" = "Condition")
    # Remove underscores
phenotype$Compound <- gsub("_", "", phenotype$Compound)
phenotype <- phenotype %>% 
      select(Strain, Compound, Normalized_Fitness)

#Pivot
phenotype2 <- phenotype %>% 
  pivot_wider(names_from = Compound_FullName, values_from = Normalized_Fitness)

# Create a second dataframe with the rownames
rownames_df <- phenotype2[[1]]  # Store the rownames in a separate vector

# Remove the first column from the original dataframe
df_matrix <- phenotype2[,-1]

# Ensure that all values are numeric and not lists
df_matrix <- as.data.frame(lapply(df_matrix, function(x) as.numeric(unlist(x))), check.names = FALSE)

# Convert the dataframe to a matrix
matrix_df <- as.matrix(df_matrix)

# Assign the rownames from the second dataframe to the matrix
rownames(matrix_df) <- rownames_df

#Annotate Strains
phylo <- read_xlsx('Additional File 1 - Supplemental Tables.xlsx', sheet = 1, skip = 2) %>% 
  select(Strain, `Phylogenetic Clade`)

annot_row <- phylo
annot_row <- annot_row %>% 
  column_to_rownames("Strain") %>% 
  as.data.frame()

ann_colors = list(
    'Phylogenetic Clade'= c(
      "Wine/European" = "#E07082",
      "Pacific West Coast Wine" = "#61d04f",
      "Transpacific Oak" = "#8F6249",
      "Mosaic" = "#c7c7c7",
      "CHN" = "#8c32bb",
      "Beer 1/Mixed Origin" = "#ffbe6fff",
      "Ecuadorian" = "#85C9F1",
      "Far East Asia" = "#6666FF",
      "Mediterranean Oak" = "#b28a6e",
      "Prise de Mousse/Vin7" = "#B24A5A",
      "Laboratory" = "#FFB3FF"),
    'Drug Category' = c(
      "Amino Acid & Protein Stability" = "#8DD3C7",
      "Cell Wall Stability" = "#FFFFB3",
      "Lipid Synthesis" = "#FB8072",
      "Membrane Integrity & Function" = "#80B1D3",
      "Iron Homeostasis" = "#FDB462",
      "Microtubule Dynamics" = "#B3DE69",
      "Respiration & Mitochondrial Dynamics" = "#FCCDE5",
      "Nucleic Acid Metabolism & DNA Damage" = "#BEBADA",
      "Oxidative Stress" = "#BC80BD",
      "Signal Transduction" = "#CCEBC5"
    ))

drugs <- read_xlsx('Additional File 1 - Supplemental Tables.xlsx', sheet = 2, skip = 2)
#Drug Category
drugs <- drugs %>% 
  select(`Drug Category`, `Drug/Compound Name`)
  
#Drug Use
annot_col <- drugs %>% 
  rename("Compound" = `Drug/Compound Name`) %>% 
  drop_na()

annot_col <- annot_col %>% 
  column_to_rownames("Compound") %>% 
  as.data.frame()

#Z normalize
heatmap_z <- t(scale(t(matrix_df)))  # Scale each row
pheatmap(heatmap_z), 
         color = colorRampPalette(c("red", "white", "blue"))(100), 
         annotation_row = annot_row,
         kmeans_k = 5,
         annotation_colors = ann_colors,
         annotation_col = annot_col,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = FALSE,
         fontsize = 12,
         fontsize_row = 1,
         fontsize_col = 17,
         width = 18,
         height = 12)

p <- pheatmap(heatmap_z, 
              color = colorRampPalette(c("red", "white", "blue"))(100), 
              fontsize_row = 0, 
              annotation_row = annot_row, 
              annotation_colors = ann_colors,
              annotation_col = annot_col,
              cluster_rows = TRUE, 
              cluster_cols = TRUE, 
              show_rownames = FALSE,  
              width = 11,
              height = 7)
```


